Exploiting First-Class Arrays in Fortran for Accelerator Programming

### 2010-12-12 opencl/papers/intro.tex
more

At this point in time it is not really possible to write once and
run efficiently on the wide variety of computer platforms we have
available.  For some classes of applications, we believe that this goal
is possible using language constructs already present in a popular
mainstream scientific programming language -- Fortran.

@@

This paper addresses this issue by examining features in Fortran that
allow programmers to express algorithms at a very high level but yet
can be transformed by a compiler to run efficiently on a wide variety
of platforms.  In particular we consider nodes of GPUs.

We demonstrate that the array syntax of Fortran maps surprisingly well
onto GPUs when transformed to OpenCL kernels.  These Fortran language
features include pure and elemental functions and array constructs
like where and shift.  In addition we add a few functions that enable
a program to be deployed on machines with a hierarchy of processing
elements, such as nodes employing GPU acceleration, \emph{without
  requiring explicit declaration of parallelism within the program.}
In addition the program uses entirely standard Fortran so it can be on
a single core without concurrency.

@@

%% NEW - CER - copied from proposal
An initial study was made for an important procedure in PAGOSA,
production code at LANL completely written in data-parallel
Fortran.  We investigated automatically transforming this code
to run on LANL's Petaflop Roadrunner computer (a hybrid mixture of Opterons
and IBM Cell processors).  We demonstrated that a source-to-source compiler
can automatically vectorize and parallelize (both within the Cell processor
and across nodes) a small section of this code.  Preliminary results showed a
12 times performance gain on the Cell processor over a traditional single core
processor.  A factor of 9 is gained from vectorizing and parallelizing the
code to run on the Cell and the rest is gained from automatically overlapping
inter-node communication with computation on the Cell.

@@

%% NEW - CER
Fortran \footnote{Originally named FORTRAN for FORmula TRANslation language.}
is the oldest high-level programming language, and was developed to facilitate
the translation of math formulae into machine code. Fortran was the first
major language to use a compiler to translate from a high-level program
representation to assembly language. Due to its age, it carries certain arcane
baggage that later programming languages have evolved away from. As a result,
Fortran has fallen into disfavor in certain programming circles.

%% NEW - CER
However, with the introduction of Fortran 90 (and later revisions to the
standard), Fortran is a truly modern programming language. Much of the arcane
baggage associated with it is actually no longer part of the language used by
modern programmers, and exists in a deprecated form simply to support legacy
codes. It is now modular and has many object-oriented features. Furthermore,
Fortran should be of interest to those studying parallel programming because
of its functional and data parallel constructs and because of the coarray
notation introduced after Fortran 2003. Unlike languages like C and C++, which
are considered to be more modern, Fortran has become a truly parallel language
with features added to recent language standards.

>>>>

At this point in time it is not really possible to write once and
run efficiently on the wide variety of computer platforms we have
available.  For some classes of applications, we believe that this goal
is possible using language constructs already present in a popular
mainstream scientific programming language -- Fortran.
A common, long-standing tongue-in-cheek response to new language developments
in the scientific and high performance computing community is that a new language
will arise to answer the needs of new systems, and it will be called Fortran.  
We believe that the work presented in this paper validates that notion -- 
we need a new language to work with, and that language is Fortran.

@@

This paper addresses the accelerator programmign problem by examining
features in Fortran that allow programmers to express algorithms at a
very high level that can be easily transformed by a compiler to run
efficiently on a wide variety of platforms.  In particular we consider
computers based on GPUs and related accelerator processors.

We demonstrate that the array syntax of Fortran maps surprisingly well
onto GPUs when transformed to OpenCL kernels.  These Fortran language
features include pure and elemental functions and array constructs
like where and shift.  In addition we add a few functions that enable
a program to be deployed on machines with a hierarchy of processing
elements, such as nodes employing GPU acceleration, \emph{without
  requiring explicit declaration of parallelism within the program.}
In addition the program uses entirely standard Fortran so it can be on
a single core without concurrency.  This work also is applicable
to vendor-specific languages similar to OpenCL such as the NVidia
CUDA language.

@@

An initial study was made for an important procedure in PAGOSA, a
non-research, production-grade code at LANL completely written in
data-parallel Fortran.  We investigated automatically transforming
this code to run on LANL's Petaflop Roadrunner computer (a hybrid
mixture of AMD Opterons and IBM Cell processors).  We demonstrated that a
source-to-source compiler can automatically vectorize and parallelize
(both within the Cell processor and across nodes) a small section of
this code.  Preliminary results showed a 12 times performance gain on
the Cell processor over a traditional single core processor.  A factor
of 9 is gained from vectorizing and parallelizing the code to run on
the Cell and the rest is gained from automatically overlapping
inter-node communication with computation on the Cell.

@@

Fortran (originally named FORTRAN, the FORmula TRANslation
language\footnote{The capitalization convention for referring the
  language was changed in 1990 to \emph{Fortran}}) is the oldest
high-level programming language in continuous use since its
introduction, and was developed to facilitate the translation of math
formulae into machine code. Fortran was the first major language to
use a compiler to translate from a high-level program representation
to assembly language. Due to its age, it carries certain arcane
baggage that later programming languages have evolved away from. As a
result, Fortran has fallen into disfavor in certain programming
circles.  Modern versions of the language standardized in 1990, 1995, 2003, and 2008
have removed much of this legacy baggage, but these changes are not
widely known.  Modern Fortran exhibits features that are similar to 
other modern programming languages, and does not mandate legacy features
from decades old, deprecated versions of the language.

With the introduction of Fortran 90 (and later revisions to the
standard), Fortran became a truly modern programming language. Much of
the arcane baggage associated with it is actually no longer part of
the language used by modern programmers, and exists in a deprecated
form simply to support legacy codes. It is now modular and has many
object-oriented features.  It now includes a type system in which rich
first-class array data types and corresponding syntax are part of the
language, something that languages like C continue to lack.
Furthermore, Fortran should be of interest to those studying parallel
programming because of its functional and data parallel constructs and
because of the coarray notation introduced in Fortran 2008. Unlike
languages like C and C++, which are considered to be more modern,
Fortran has become a truly parallel language with features added to
recent language standards.
